### 2020-07-27 Catuscia - submitted/07-discussion.tex
da dropbox

Our encoding increases the expressivity of  RS  concerning:
the behaviour of the context, the possibility of alternative behaviour of
mutated entities and the communication between two different reaction systems.
It is important to note that our encoding guarantees that from each state, 
in the  c\CNA \ transition system, only one transition comes out, 
as the dynamics is totally deterministic.

@@

\subsection{Mutating entities} 
In RS, when an entity is present, it can potentially be involved in each reaction where it is required.
With a few more lines of code, in c\CNA~it is possible to describe the behaviour of a mutation of 
an entity, in a way that the mutated version of the entity can take 
part to only a subset of the rules requiring  the  \emph{normal version} of the entity.
For example, let us assume that entity $s1$ is consumed by reactions $a1$ and $a2$.
Reaction $a1$ produces $s1$ if $s2$ is present, otherwise $a1$ produces a mutated version of $s1$, 
say $s1'$.
When $s1'$ is produced, reaction $a1$ behaves in the same way as if $s1$ would be absent, whereas $a2$
recognises the presence of $s1'$ and behaves in the same way as if $s1$ would be present.
Technically, in both cases it is enough to add one more non 
deterministic choice in the code of $P_{a1}$ and $P_{a2}$.

\subsection{Communicating reaction systems} 
We sketch how it is possible to program two RS encodings, in a way that 
the entities that usually come from context of one RS will be provided instead from the other RS.

@@

\begin{table}[t]
\centering
\begin{tabular}{|c|c|}
\hline
$rs1$  & $rs2$ \\
\hline
%$r1_1=(s1,\ , s2)$ & $r2_1=(s3,\  ,s1)$\\
$a_1=(s,\ , x)$ & $a_2=(y,\  ,s)$\\
%$r1_2=(\ , s2,s1)$ & $r2_2=(s1,\ ,s3)$\\
\hline 
\end{tabular}
\caption{The two reaction systems $rs1$ and $rs2$.}
\label{tab:rules}
\end{table}
 \noindent
Now, we set our example such that the two contexts, for  $rs1$ and $rs2$, do not provide any entities.
We also assume that entity $s$ in $rs1$ is provided by $rs2$, as $rs2$ produces a quantity of $s$ that is enough for $rs1$ and $rs2$.
For technical reasons, we can not use the same name for  $s$ in both the two RSs, then we use the name $ss$ in $rs2$.
%
We need to modify our translation technique to suite this new setting. 
As we do not model contexts, we introduce  \emph{dummy} channel names $dx$ and $dss$ to model the case $x$ and $ss$ are not produced.
%absence of entities. 
Also, thanks to the simplicity of the example, we can leave out the use of the   $p_i$ channels. This streamlining does not affect the programming technique we propose to make two RSs communicate.
%Thus, there will be no $Cxt$ processes and we manage the code in a way that $rs1$ will receive the entity $s1$ from $rs2$. 
%Please also note that, for our simple example, we do not have anymore to specify the all steps
%of a interactive system, as the communication mechanism from $rs2$ to $rs1$ will generate an infinite
%behaviour.
%\noindent
%First, we translate the rules in $rs1$:
First, we translate the reaction in $rs1$:
\[
\begin{array}{lcl}
%\llbracket r1_1\rrbracket \defeq P_{r1_1}&\defeq &\startchain{\silent}\chainedlink{s1}{\noact}\chainedlink{\noact}{s1}\chainedlink{\tilde{s2}}{\noact}\chainedlink{\noact}{\tilde{s2}}\chainend{r1_2}.P_{r1_1} +
%\startchain{\silent}\chainedlink{\overline{s1}}{\noact}\chainedlink{\noact}{\overline{s1}}\chainend{r1_2}.P_{r1_1}\\
\llbracket a_1\rrbracket \defeq P_{a_1}&\defeq &
\startchain{\silent}\chainedlink{s_i}{{\color{blue}\noact}}\chainedlink{{\color{blue}\noact}}{s_o}\chainedlink{\tilde{x}_i}{{\color{gray}\noact}}\chainedlink{{\color{gray}\noact}}{\tilde{x}_o}\chainend{a_2}.P_{a_1} +
\startchain{\silent}\chainedlink{\overline{s}_i}{\noact}\chainedlink{\noact}{\overline{s}_o}    \chainedlink{dx_i}{\noact}\chainedlink{\noact}{dx_o}\chainend{a_2}.P_{a_1}\\
%\end{array}
%\]
%\\
%
%\begin{array}{lcl}
%\llbracket r1_2\rrbracket \defeq P_{r1_2}&\defeq &\startchain{r1_2}\chainedlink{\overline{s2}}{\noact}\chainedlink{\noact}{\overline{s2}}\chainedlink{\tilde{s1}}{\noact}\chainedlink{\noact}{\tilde{s1}}\chainend{r2_1}.P_{r2_1} +
%\startchain{r1_2}\chainedlink{s2}{\noact}\chainedlink{\noact}{s2}\chainend{r2_1}.P_{r2_1}\\
\end{array}
\]

@@

As we have briefly sketched, our model of two 
\emph{communicating reaction systems} 
can enable the study of the behaviour of one RS
in relation to another one.
% In Section~\ref{ex:lactose} describing  the behaviour of the \emph{lac} operon,
% the two entities lactose and glucose are controlled 
% non deterministically by the context.
% In our framework,
% instead, by exploiting the expressivity of c\CNA, 
% the \emph{lac} operon system can be connected with the two systems 
% producing the lactose and the glucose.
% This way, the presence of these two entities in the \emph{lac} operon system 
% could be regulated by realistic mechanisms.
Thus, the products of the reactions of one RS can become the input for
another one. This could allow for a modular approach to modeling 
complex systems, by composing different Reaction Systems.

>>>>

Our encoding increases the expressivity of  RS  concerning:
the behaviour of the context, the possibility of alternative behaviour of
mutated entities, the communication between two different reaction systems.
It is important to note that our encoding guarantees that from each state, 
in the  c\CNA \ transition system, only one transition comes out, 
as the dynamics is totally deterministic.

@@

\subsection{Mutating entities} 
In RS, when an entity is present, it can potentially be involved in each reactions where it is required.
With a few more lines of code, in c\CNA~it is possible to describe the behaviour of a mutation of 
an entity, in a way that the mutated version of the entity can take 
part to only a subset of the rules requiring  the  \emph{normal version} of the entity.
For example, let us assume that entity $s1$ is consumed by reactions $a1$ and $a2$.
Reaction $a1$ produces also $s1$ if $s2$ is present, otherwise $a1$ produces a mutated version of $s1$, 
say $s1'$.
When $s1'$ is produced, reaction $a2$ behaves in the same way as if $s1$ would be absent, whereas $a2$
recognises the presence of $s1'$ and behaves in the same way as if $s1$ would be present.
Technically, in both cases it is enough to add one more non 
deterministic choice in the code of $P_{a1}$ and $P_{a2}$.

\subsection{Communicating reaction systems} 
We sketch how it is possible to program two RSs encodings, in a way that 
the entities that usually come from context of one RS will be provided instead from the other RS.

@@

\begin{table}[t]
\centering
\begin{tabular}{|c|c|}
\hline
$rs1$  & $rs2$ \\
\hline
%$r1_1=(s1,\ , s2)$ & $r2_1=(s3,\  ,s1)$\\
$a_1=(s,\ , x)$ & $a_2=(y,\  ,s)$\\
%$r1_2=(\ , s2,s1)$ & $r2_2=(s1,\ ,s3)$\\
\hline 
\end{tabular}
\caption{The two reaction systems $rs1$ and $rs2$.}
\label{tab:rules}
\end{table}
 \noindent
Now, we set our example such that the two contexts, for  $rs1$ and $rs2$, do not provide any entities.
We also assume that entity $s$ in $rs1$ is provided by $rs2$, as $rs2$ produces a quantity of $s$ that is enough for $rs1$ and $rs2$.
For technical reasons, we can not use the same name for  $s$ in both the two RSs, then we use the name $ss$ in $rs2$.
%
We need to modify our translation technique to suite this new setting. 
As we do not model contexts, we introduce  \emph{dummy} channel names $dx$ and $dss$ to model the absence of entities. Also, thanks to the simplicity of the example, we can leave out the use of the   $p_i$ channels. This streamlining does not affect the programming technique we propose to make two RSs communicate.
%Thus, there will be no $Cxt$ processes and we manage the code in a way that $rs1$ will receive the entity $s1$ from $rs2$. 
%Please also note that, for our simple example, we do not have anymore to specify the all steps
%of a interactive system, as the communication mechanism from $rs2$ to $rs1$ will generate an infinite
%behaviour.
%\noindent
%First, we translate the rules in $rs1$:
First, we translate the reaction in $rs1$:
\[
\begin{array}{lcl}
%\llbracket r1_1\rrbracket \defeq P_{r1_1}&\defeq &\startchain{\silent}\chainedlink{s1}{\noact}\chainedlink{\noact}{s1}\chainedlink{\tilde{s2}}{\noact}\chainedlink{\noact}{\tilde{s2}}\chainend{r1_2}.P_{r1_1} +
%\startchain{\silent}\chainedlink{\overline{s1}}{\noact}\chainedlink{\noact}{\overline{s1}}\chainend{r1_2}.P_{r1_1}\\
\llbracket a_1\rrbracket \defeq P_{a_1}&\defeq &
\startchain{\silent}\chainedlink{s_i}{{\color{blue}\noact}}\chainedlink{{\color{blue}\noact}}{s_o}\chainedlink{\tilde{x}_i}{{\color{gray}\noact}}\chainedlink{{\color{gray}\noact}}{\tilde{x}_o}\chainend{a_2}.P_{a_1} +
\startchain{\silent}\chainedlink{\overline{s}_i}{\noact}\chainedlink{\noact}{\overline{s}_o}    \chainedlink{dx_i}{\noact}\chainedlink{\noact}{dx_o}\chainend{a_2}.P_{a_1}\\
%\end{array}
%\]
%\\
%
%\begin{array}{lcl}
%\llbracket r1_2\rrbracket \defeq P_{r1_2}&\defeq &\startchain{r1_2}\chainedlink{\overline{s2}}{\noact}\chainedlink{\noact}{\overline{s2}}\chainedlink{\tilde{s1}}{\noact}\chainedlink{\noact}{\tilde{s1}}\chainend{r2_1}.P_{r2_1} +
%\startchain{r1_2}\chainedlink{s2}{\noact}\chainedlink{\noact}{s2}\chainend{r2_1}.P_{r2_1}\\
\end{array}
\]

@@

As we have briefly sketched, our model of two \emph{communicating reaction systems} 
can enable the study of the behaviour of one RS
in relation to another one.
In Example~\ref{ex:twoRS} describing  the behaviour of the \emph{lac} operon,
the two entities lactose and glucose are controlled 
non deterministically by the context.
In our framework,
instead, by exploiting the expressivity of c\CNA, 
the \emph{lac} operon system has been connected with the two systems 
producing the lactose and the glucose.
This way, the presence of these two entities in the \emph{lac} operon system 
can be regulated by realistic mechanisms.
%which can be modified to represent pathological states as well.

% Again, we stress the fact the verbosity of the {\tt link}-calculus code 
%should not be
% seen as a drawback, as it undergoes to automatic code translation.

### 2020-07-27 Catuscia - submitted/08-conclusion.tex
da dropbox

In this paper we have exploited a variant of the {\tt link}-calculus
where prefixes are link chains and no more single links. This variant was already briefly discussed by the end of~\cite{BBB17}.
This variant allowed us to define 
%shown 
an elegant embedding of reaction systems,
an emerging formalism to model computationally biochemical systems.
%into multilink processes.  
This translation shows several benefits.
For instance, the context behaviour can also be expressed recursively;
entity mutations can be expressed easily; reaction systems can
communicate between them.

We believe that our embedding can contribute to extend the applications
of reaction systems to diverse fields of computer science, and life
sciences.
As we have already mentioned, the evolution of each process resulting from our embedding 
is deterministic, thus we do not have the problem of having infinitely many transitions
in the produced labelled transition system. 
In any case, we can exploit the implementation of  the symbolic semantics of the {\tt link}-calculus~\cite{BrodoO17} that is available at~\cite{tool}.

As future work, we plan to implement a prototype of our framework,
with an automatic translation from RSs to the {\tt link}-calculus. 
We believe that our work can also help to extend the framework
of RSs towards a model which can improve the communication
between different RSs. We also believe that our work can make 
possible to investigate how to apply formal techniques to prove 
properties of the
modeled systems 
\cite{CFHOT15,OCHF16,BBGLBH2017,FOP15,CFOP10}.

\medskip

\noindent
{\small {\bf Acknowledgments}
We thank the anonymous reviewers for their detailed and very useful 
criticisms and recommendations that helped us to improve our paper.}

>>>>

In this paper we have introduced a variant of the {\tt link}-calculus
where prefixes are link chains and no more single links, as it was 
briefly described in the future work section in~\cite{BBB17}.
This variant allowed us to define 
%shown 
an elegant embedding of reaction systems,
an emerging formalism to model computationally biochemical systems.
%into multilink processes.  
This translation shows several benefits.
For instance, the context behaviour can also be expressed recursively. 
Entity mutations can be expressed easily. Reaction systems can
communicate between them.

We believe that our embedding can contribute to extend the applications
of reaction systems to diverse fields of computer science, and life
sciences.
As we have already mentioned, the evolution of each process resulting from our embedding 
is deterministic, thus we do not have the problem of having infinitely many transitions
in the produced labelled transition system. In any case, we can exploit the implementation of  the symbolic semantics of the {\tt link}-calculus~\cite{BrodoO17} that can be found in~\cite{tool}.

As future work, we plan to implement a prototype of our framework,
with an automatic translation from RSs to  {\tt link}-calculus. 
We believe that our work can also help to extend the framework
of RSs towards a model which can improve the communication
between different RSs. We also believe that our work can make 
possible to investigate how to apply formal techniques to prove 
properties of the
modeled systems \cite{CFHOT15,OCHF16,BBGLBH2017}.

### 2020-07-27 fragments/01-introduction.tex
da dropbox

As we have remarked, in our translation, Reaction Systems 
get the ability to interact between them in a synchronized manner. 
This interaction is not foreseen in
the basic RS framework, as it can only happen with the context.
By exploiting recursion, the kind of interactions which can be 
defined can be complex and expressive.
Example~\ref{ex:twoRS} and more in general the discussion in 
Section~\ref{sec:discussion} show that 
the interaction between RSs can help to model new scenarios.

>>>>

As we have remarked, in our translation, Reaction Systems 
get the ability to interact between them in a synchronized manner. 
This interaction is not foreseen in
the basic RS framework, as it can only happen with the context.
By exploiting recursion, the kind of interactions which can be 
defined can be complex and expressive.
Example~\ref{ex:twoRS} and more in general the discussion in 
Section~\ref{sec:discussion} show that 
the interaction between RSs can help to model new scenarios.
\paragraph{Related work}
We are not aware of any inference rules for the RS; it seems not the case
that a deterministic behavior, as the one of the RS, once the initial state is 
fixed, could be defined by inference rules that classically are applied 
to define non-deterministic transition systems.
The reversible computation paradigm for RS extends the RS framework  by
allowing  backward computations as well as forward computations; for this 
goal a set of inference rules for rewriting logic has been 
defined in~\cite{10.1007/978-3-319-73359-3_3}
to exactly keep trace of those elements that  dissolve in the next 
computation step.
